### 2024-06-19 basic-statistics.qmd
20240618 - tweaks

A two-by-two confusion matrix for null-hypothesis significance testing is in @fig-nhstConfusionMatrix.
There are four general possibilities:

1. We can (correctly) reject the null hypothesis when it is in fact false ($1 - \beta$).
This is a true positive.
For instance, we may correctly determine that Quarterbacks have longer careers than Running Backs.
1. We can (correctly) fail to reject the null hypothesis when it is in fact true ($1 - \alpha$).
This is a true negative.
For instance, we may correctly determine that Quarterbacks do not have longer careers than Running Backs.
1. We can (incorrectly) reject the null hypothesis when it is in fact true ($\alpha$).
This is a false positive.
When performing a null hypothesis testing, a false positive is known as a Type I error.
For instance, we may incorrectly determine that Quarterbacks have longer careers than Running Backs when, in fact, Quarterbacks and Running Backs do not differ in their career length.
1. We can (incorrectly) fail to reject the null hypothesis when it is in fact false ($\beta$).
This is a false positive.
When performing a null hypothesis testing, a false negative is known as a Type II error.
For instance, we may incorrectly determine that Quarterbacks and Running Backs do not differ in their career length when, in fact, Quarterbacks have longer careers than Running Backs.

@@

A plot of statistical power is in @fig-nhst.
In statistics, *power* is the probability of detecting an effect, if, in fact, the effect exists.
Otherwise said, power is the probability of rejecting the null hypothesis, if, in fact, the null hypothesis is false.
Power is influenced by several variables:

- the sample size (*N*): the larger the *N*, the greater the power
- the [effect size](sec-practicalSignificance): the larger the effect, the greater the power
    - for group comparisons:
        - larger between-group variance, and
        - smaller within-group variance (i.e., strong measurement precision)
- the alpha level (typically .05): the higher the alpha level, the greater the power; however, the higher we set the alpha level, the higher the likelihood of Type I errors (false positives)
- one- versus two-tailed tests: one-tailed tests have higher power than two-tailed tests
- within-subject versus between-subject comparisons: within-subject designs tend to have greater power than between-subject designs

@@

*Practical significance* deals with how important the effect/association/difference is.
It is based on the magnitude of the effect, called the *effect size*.
Effect size can be quantified in various ways including:

>>>>

There are four general possibilities of decision making outcomes when performing null-hypothesis significance testing:

1. We (correctly) reject the null hypothesis when it is in fact false ($1 - \beta$).
This is a true positive.
For instance, we may correctly determine that Quarterbacks have longer careers than Running Backs.
1. We (correctly) fail to reject the null hypothesis when it is in fact true ($1 - \alpha$).
This is a true negative.
For instance, we may correctly determine that Quarterbacks do not have longer careers than Running Backs.
1. We (incorrectly) reject the null hypothesis when it is in fact true ($\alpha$).
This is a false positive.
When performing null hypothesis testing, a false positive is known as a Type I error.
For instance, we may incorrectly determine that Quarterbacks have longer careers than Running Backs when, in fact, Quarterbacks and Running Backs do not differ in their career length.
1. We (incorrectly) fail to reject the null hypothesis when it is in fact false ($\beta$).
This is a false positive.
When performing null hypothesis testing, a false negative is known as a Type II error.
For instance, we may incorrectly determine that Quarterbacks and Running Backs do not differ in their career length when, in fact, Quarterbacks have longer careers than Running Backs.

A two-by-two confusion matrix for null-hypothesis significance testing is in @fig-nhstConfusionMatrix.

@@

In statistics, *power* is the probability of detecting an effect, if, in fact, the effect exists.
Otherwise said, power is the probability of rejecting the null hypothesis, if, in fact, the null hypothesis is false.
Power is influenced by several variables:

- the sample size (*N*): the larger the *N*, the greater the power
- the [effect size](#sec-practicalSignificance): the larger the effect, the greater the power
    - for group comparisons:
        - larger between-group variance, and
        - smaller within-group variance (i.e., strong measurement precision)
- the alpha level (typically .05): the higher the alpha level, the greater the power; however, the higher we set the alpha level, the higher the likelihood of Type I errors (false positives)
- one- versus two-tailed tests: one-tailed tests have higher power than two-tailed tests
- within-subject versus between-subject comparisons: within-subject designs tend to have greater power than between-subject designs

A plot of statistical power is in @fig-nhst.

@@

*Practical significance* deals with how large or important the effect/association/difference is.
It is based on the magnitude of the effect, called the *effect size*.
Effect size can be quantified in various ways including:
